---
title: "Restaurant Booking Analysis with Bootstrapping"
author: "Daniele Melotti"
date: "Dec 2023"
output: pdf_document
---

# 1. Data Exploration

## Load the data and convert _datetime_ into a POSIXlt date-time format

We start by loading and taking a peek at the dataset:
```{r}
bookings <- read.table("../data/first_bookings_datetime_sample.txt",
                       header=TRUE) # loading the data
str(bookings$datetime) # understanding what data type are we dealing with
```

As we can see, our data is contained in a character vector, which is not ideal for performing statistical operations. Therefore, we will convert the content of _datetime_ to a POSIXlt date-time format and extract the components that interest us.

## Calculate the time of each booking in minutes since the start of the day. Store the result in a new variable - _minday_

Since we know that EZTABLE would like to start a promotion for new members to make their bookings earlier in the day, we focus on extracting the hours and minutes after converting _datetime_ to a POSIXlt object.
```{r}
# Converting datetime to a POSIXlt object, extracting hours and minutes
hours <- as.POSIXlt(bookings$datetime, format="%m/%d/%Y %H:%M")$hour 
mins <- as.POSIXlt(bookings$datetime, format="%m/%d/%Y %H:%M")$min
```

Afterwards, we calculate the time of each booking in minutes since the start of the day. So, we multiply the hours by 60 and add the minutes, resulting in a single numerical value representing the minute of the day:
```{r}
minday <- hours * 60 + mins
```

## Plot the density of _minday_

Now, we shall see the density of _minday_ so as to get an idea about how are the values of this vector distributed:
```{r}
plot(density(minday), main="Minute of the day of first ever booking",
     col="cornflowerblue", lwd=2)
```

We see a distribution that is similar to a bimodal, with a peek of booking time around 700 minutes and another around 1200 minutes, which likely represent booking requests before lunch and dinner time. Next, we are going to conduct an analysis of the mean booking time.

# 2. Mean Booking Times - Bootstrapping

## Calculate the mean booking time and its standard error.

We shall start by computing the mean booking time and its standard error. The standard error is computed by dividing the standard deviation of the booking time by the square root of the total number of bookings:
```{r}
minday_mean <- mean(minday) # mean
minday_mean
minday_se <- sd(minday)/sqrt(length(minday)) # standard error
minday_se
```

The mean booking time corresponds to 942.4964 minutes. To make the time quickly readable and familiar, we use a custom function, located in the _functions.R_ script:
```{r}
source(file = "../R/functions.R")
mins_to_time(minday_mean)
```

So, we know that the time at which a first booking is made on average is 15:42.

## Develop a 95% confidence interval for the mean booking time.

Recall that the data we have available is only a sample; now that we know mean and standard error, we are able to build a confidence interval and get a clearer idea about the true average booking time:
```{r}
# 95% CI:
lower_bound <- minday_mean - 1.96 * minday_se
lower_bound
upper_bound <- minday_mean + 1.96 * minday_se
upper_bound
```

The interval we just built suggests that we can be 95% confident that the true average booking time is between 941.2308 and 943.2308 minutes, or that it is included between 15:41 and 15:44.
```{r}
mins_to_time(lower_bound)
mins_to_time(upper_bound)
```

## Implement a bootstrap approach to generate 2000 new samples from the original dataset.

Next, we are going to employ bootstrapping. This is a technique that resamples a dataset to create several simulated samples. To implement bootstrapping, we use the `replicate()` function, paired with `sample()`. Let's generate 2000 new samples from the original sample data:
```{r}
# Setting seed
set.seed(100)

# Creating 2000 new samples (bootstrapped samples) from the starting data
new_samples <- replicate(n = 2000,
                         sample(minday, length(minday), replace = TRUE))
```

## Visualize the distribution of the means from the bootstrapped samples.

We might be interested in seeing how the 2000 resampled datasets look like in comparison with the starting sample. We can display them on the same plot. We achieve this by applying the custom function `plot_resamples()`, included in the `functions.R` script. This function plots the density of a sample and then applies a function on the sample; if paired with `apply()`, it allows us to plot each samples' density and compute their means.
```{r}
# Create an empty plotting space
plot(density(minday), lwd=0, ylim=c(0, 0.004), xlab = "Minutes", ylab = "",
     main = "Original sample vs. Bootstrapped samples with Mean indicators")

# Plot and get means of all bootstrapped samples
sample_means <- apply(new_samples, 2,
                      FUN = function(x) plot_resamples(x, mean))

# Add starting sample density to the plot
lines(density(minday), lwd = 2)

# add vertical lines for the means of each the original sample and each new sample 
abline(v = sample_means, col = rgb(0.0, 0.4, 0.0, 0.05))
abline(v = minday_mean, lwd = 1.5, col = "red")
```

We notice that the bootstrapped samples' densities are all very close to the original sample. The same goes for the bootstrapped means, the confidence interval is certainly very narrow. Now, let's visualize the density of the 2000 bootstrapped means. We can take a look at how they compare with the original mean value from the starting sample.
```{r}
plot(density(sample_means), lwd = 2, col = "cornflowerblue",
     main = "Density of Bootstrapped samples with Original mean indicator")
abline(v = minday_mean, lwd = 2, col = "tomato")
```

We see that the bootstrapped means do not vary largely from the mean of the original sample. Most values are included between 941 and 944 minutes, with only a few means being outside of this interval.

## Calculate and interpret the 95% confidence interval of the bootstrapped means.

We know what the 95% confidence interval is for the original sample's mean, now we can build a new interval for the resampled means. We can build this interval using the `quantile()` function: 
```{r}
resampled_means_CI <- quantile(sample_means, probs = c(0.025, 0.975))
resampled_means_CI
```

The 95% CI of the bootstrapped means ranges from 942.4681 to 942.5221 (or from 15:42 to 15:43).
```{r}
mins_to_time(resampled_means_CI)
```

This means that we are 95% confident that the true average booking time falls between 15:41 and 15:44. With this confidence interval, we complete the analysis of the average booking time.

# 3. Median Booking Times - Bootstrapping

## Determine the median booking time.

Now, we focus on the median booking time, in other words, the time of the day by which half of the customers has made their first booking.
```{r}
# Computing median booking time
minday_median <- median(minday)
minday_median

# Converting minutes to time
mins_to_time(minday_median)
```

We see that the median booking time consists of 1040 minutes, in other words, it's 17:20.

## Visualize the distribution of the medians from the bootstrapped samples.

We can display the bootstrapped medians and compare them with the original sample's median booking time. We shall use `plot_resamples()` again.
```{r}
# Create an empty plotting space
plot(density(minday), lwd=0, ylim=c(0, 0.004), xlab = "Minutes", ylab = "",
     main = "Original sample vs. Bootstrapped samples with Median indicators")

# Plot and get means of all bootstrapped samples
sample_medians <- apply(new_samples, 2,
                        FUN = function(x) plot_resamples(x, median))

# Add starting sample density to the plot
lines(density(minday), lwd = 2)

# add vertical lines for the means of each the original sample and each new sample 
abline(v = sample_medians, col = rgb(0.0, 0.4, 0.0, 0.05))
abline(v = minday_median, lwd = 1.5, col = "red")
```

Interestingly, we see that the bootstrapped medians are close to the original sample's median, however, there are gaps in between these values. This is due to the fact that all the bootstrapped median values are represented by one of these:
```{r}
sort(unique(sample_medians))
```

Now, let's visualize the density of the 2000 bootstrapped medians:
```{r}
plot(density(sample_medians), lwd = 2, col = "cornflowerblue",
     main = "Density of Bootstrapped samples with Original median indicator")
abline(v = minday_median, lwd = 2, col = "tomato")
```

As we can see, the bootstrapped medians are distributed quite widely and non-normally. This makes it hard to use them for inference. 

## Estimate and interpret the 95% confidence interval of the bootstrapped medians.

Let's compute the 95% CI:
```{r}
resampled_median_CI <- quantile(sample_medians, probs = c(0.025, 0.975))
resampled_median_CI
```

The confidence interval shows that we can be 95% confident that the true median booking time is included between minutes 1020 and 1050, else:
```{r}
mins_to_time(resampled_median_CI)
```

Between 17:00 and 17:30. This is a window of 30 minutes, which is quite large.

# 4. Business Implications

## Insights
summary table: original samples vs resampled; means CIs...

## Business Strategy Recommendations


