---
title: "Restaurant Booking Analysis with Bootstrapping"
author: "Daniele Melotti"
date: "Dec 2023"
output: pdf_document
---

# 1. Data Exploration

## Load the data and convert _datetime_ into a POSIXlt date-time format

We start by loading and taking a peek at the dataset:
```{r}
bookings <- read.table("../data/first_bookings_datetime_sample.txt",
                       header=TRUE) # loading the data
str(bookings$datetime) # understanding what data type are we dealing with
```

As we can see, we are dealing with a character vector, which is not ideal for performing statistical operations. Therefore, we will convert the content of _datetime_ to a POSIXlt date-time format and extract the component that interest us.

## Calculate the time of each booking in minutes since the start of the day. Store the result in a new variable - _minday_

Since we know that EZTABLE would like to start a promotion for new members to make their bookings earlier in the day, we focus on extracting the hours and minutes after converting _datetime_ to a POSIXlt object.
```{r}
# Converting datetime to a POSIXlt object, extracting hours and minutes
hours <- as.POSIXlt(bookings$datetime, format="%m/%d/%Y %H:%M")$hour 
mins <- as.POSIXlt(bookings$datetime, format="%m/%d/%Y %H:%M")$min
```

Afterwards, we calculate the time of each booking in minutes since the start of the day. So, we multiply the hours by 60 and add the minutes, resulting in a single numerical value representing the minute of the day:
```{r}
minday <- hours*60 + mins
```

## Plot the density of _minday_

Now, we shall see the density of _minday_ so as to get an idea about how are the values of this vector distributed:
```{r}
plot(density(minday), main="Minute of the day of first ever booking",
     col="cornflowerblue", lwd=2)
setwd("/Users/danielemelotti/Library/CloudStorage/GoogleDrive-daniele.melotti@gapp.nthu.edu.tw/My Drive/5. REPOS/restaurant-booking-bootstrap")
```

We see a distribution that is similar to a bimodal, with a peek of booking time around 700 minutes and another around 1200 minutes, which likely represent booking requests before lunch and dinner time. 

